Verification of concurrent design patterns with data

### 2019-02-18 2018-Arch-pNets/arch-pNets.tex
sectino 4 et annexes

\noteEM{Je m'arrete la, j'\'etais suppos\'e passer apr\`es Simon... (;-)}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.9\columnwidth]{FailureTimerPNET-v4}
  \caption{The open pNET encoding the Failure Monitor architecture (\fig{schema:ArchFailure:BIP}) without the priority model}
  \label{fig:FailureTimer:pNet}
%% %
%%   \noteSBin{The updated encoding exports more primed variables in the
%%     actions, \eg $tick(z,t,z',t')$ and $fail(zone, zone', b1)$ instead
%%     of $tick(z,t,z')$ and $fail(zone, b1)$, and SVs, \eg $\dots \land
%%     (t' = t) \land (zone' = zone)$ in the guard of SV0. Shall we
%%     update the figure to include these or, rather, explain that the
%%     figure is cleaned-out for readability?}
%
\end{figure}

@@

The encoding of the interaction model is based on its representation
as a set of connectors. \noteSB{%
%
  A legitimate question is whether we actually need to do this,
  instead of listing explicitly the individual interactions, then
  associating a synchronisation vector to each one of them?  One of
  the factors is the encoding pNets$\rightarrow$SAT: is it better to
  have more or fewer SVs if the price is additional Boolean variables?
  The other factor might be the facility of maximal progress
  encoding\mdash TBC.  Traceability.  Anything else?\newline
  \hl{UPDATE: Maximal progress encoding requires this!}
%
} Indeed, as illustrated by the Failure Monitor architecture in
\fig{schema:ArchFailure:BIP}, each connector can define several
allowed interactions, depending on its hierarchical structure and the
use of synchrons and triggers.

@@

\addSB{In the general encoding, the global actions of the
  synchronisation vectors should be fresh action names.  However, for
  the properties considered in our case study, we only need to exhibit
  the actions of the holes.  Furthermore, each synchronisation vector
  involves at most one such action.  This allows us to reuse them as
  the global actions of the synchronisation vectors, simplifying the
  translation of properties to be verified (\cf \secn{arch:verif}).}

@@

Below, we assume that, as in \fig{schema:ArchFailure:BIP}, the
interaction model is defined by a set of connectors, annotated with
Boolean guards and with update assignments.  In particular, we assume
that the guards and update assignments are well-defined for any
interaction allowed by the connector.  For example, the choice
$\NameCtrl.\PortFail\ ?\ \NameCtrl.\mathit{zone} : \topinterval$ in
the update assignement
$
\NameTimer.\mathit{z} := \NameTimer.\mathit{z}
\cap \bigl(
(
\NameCtrl.\PortFail\ ?\ \NameCtrl.\mathit{zone} : \topinterval
)
+ \NameTimer.\mathit{t} \bigr)
$
associated to the connector {\NameTimer.\PortTick \trigsynch
  (\PortFail \trigsynch \NameCtrl.\PortFail)} in
\fig{schema:ArchFailure:BIP} ensures that the assignment is
well-defined independently of whether {\NameCtrl.\PortFail}
participates or not.  Let us denote by $\gamma \subset \ac(P_A)
\times \guards{V_A} \times \assigns{V_A}$ the set of connectors in the
architecture $A$ and by $P_x$ the set of ports involved in the
connector $x \in \gamma$.  Then, \addSB{the interaction model defined by $\gamma$ is 
  $\Gamma = \setdef{(a,g,u)}{(x,g,u) \in \gamma, a \in \intsem{x}}$
  and the encoding of $\Gamma$ is} $\nopri[\Gamma] \bydef{=}
\setdef{\nopri[x,g,u]}{(x,g,u) \in \gamma}$, where $\nopri[x,g,u]$ is the synchronisation vector defined by
%
\begin{multline}
  \label{eq:nopri:connector}
  \nopri[x,g,u] \bydef{=}
  %% \Tuple{p\bigl(\export[A](p),{{\color{blue}\export[A]'(p),}}b_p\bigr)}^{p \in P_x}
  \Tuple{\Setdef{p\bigl(\export[A](p),\export[A]'(p),b_p\bigr)}{p \in P_x \cap P}}^{P \in P_C^{C \in \cC} \cup \partition}
  \rightarrow
  \alpha_p
  \\
  \left[
    \bigwedge R(\tau(x))\bigl[\nicefrac{b_p}{p}\bigr]
    \land
    \bigwedge_{(x := e_x) \in u} (?x' = e_x)
    \land
    \bigwedge_{x \in \export[A](x), x \not\in u} (?x' = x)
    \right]
  \,,
\end{multline}
%% \noteSB{Ought to group ports by components in
%%   $\Tuple{p\bigl(\export[A](p),\export[A]'(p),b_p\bigr)}^{p \in P_x}$...
%%   \newline
%%   Will probably restrict interactions to one port per component
%%   throughout the paper to simplify.
%% }
%
with $b_p$ a fresh Boolean variable, $\alpha_p$ a fresh name, $\tau :
\ac(P_A) \rightarrow \ct(P_A)$ and $R : \ct(P_A) \rightarrow
\cru(P_A)$ the two mappings defined in \app{transformations} and
illustrated in \tab{connectors}, $\bigl[\nicefrac{b_p}{p}\bigr]$ is
the substitution that replaces in the expression that preceeds it
all occurrences of $p$ by $b_p$.

@@

\bigskip
In order to encode the maximal progress priority model, one has to
introduce an additional Boolean variable, for each port, for which we
have introduced one in $\nopri[A, \partition]$.  Intuitively, the
variables introduced above determine whether the original transition
labeled by the port is executed (\true), or rather the corresponding
self-loop, introduced by the encoding (\false).  The new variables
determine whether \emph{there is an original transition labelled by
  $p$ that could be executed from the same state}, \ie[,] with $b_p$
the variable introduced above for the encoding without maximal
progress, $q \goesto{p(\true, b_p)} q'$ iff $\exists q'': q
\goesto{p(b_p)} q''$ with $b_p = \true$ in $\nopri[A, \partition]$.
Then, in the SV guard, we have to check whether \emph{all ports
  leading to $p$ in the caual tree $\tau(x)$ can be fired} (see
\app{transformations}).  If so, then $p$ must be fired, \ie
$p(\false)$ must be blocked.  %% This is pretty much straightforward, but
%% care must be taken when doing this in the general case, since causal
%% interaction trees can get pretty tricky due to $\oplus$ (see
%% \app{trees}).}
%****************************************************************
%****************************************************************

@@

   {t1---cancel(false)-->t1, s1---resume(false)-->s1},   {---:hb_B-->},
   [(:hb_B=fail(b0:sva_SV0)
     /\((cancel(false))=(cancel(b1:sva_SV1:1:2)))
     /\((resume(false))=(resume(b2:sva_SV1:1:2)))
     /\(b1:sva_SV1:1:2=b2:sva_SV1:1:2)],   {}
OT4 ------------------------------------------------------------
    <t1_s1>-----resume----><t1_s1>

@@

  In particular, the first safety property \eq{prop:reset:can}:
  \begin{equation*}
    \AG \bigl(\PortFail(true) \rightarrow \EU[]{\lnot \PortFail(true)}{(\PortResume(true) \lor \PortReset)}\bigr).
  \end{equation*}
is not true, because the $OT_2$ loop on state $\ostate{11}$ stays in
state 11 where you need another $\PortFail(true)$ before enabling a
resume (or reset). This is because without priorities we can have an
activity on the dandling ``fail'' port without synchronising with the
others, even if they are available.

@@

pLTS: Timer
- Transition: t0---init-->t1, { t := 0 z := TopInt}
- Transition: t1---tick(z, t, z1)-->t1, [t<(GET_MAX z)], { t := t+1 z := z1}
- Transition: t1---cancel(true, false)-->t1, {}
- Transition: t1---cancel(true, true)-->t1, { t := 0 z := TopInt}
- Transition: t1---timeout-->t1, [t in z], { t := 0 z := TopInt}
pLTS: Failure
- Transition: s0---init-->s1, { zone := minZone,maxZone}
- Transition: s1---timeout-->s1, {}
- Transition: s1---resume(false, false)-->s1, {}
- Transition: s1---fail(zone, true, false)-->s1, {}
- Transition: s1---fail(zone, true, true)-->s2, {}
- Transition: s2---fail(zone, false, false)-->s2, {}
- Transition: s2---resume(true, false)-->s2, {}
- Transition: s2---resume(true, true)-->s1, {}
- Transition: s2---timeout-->s3, {}
- Transition: s3---fail(zone, false, false)-->s3, {}
- Transition: s3---resume(false, false)-->s3, {}
- Transition: s3---reset-->s1, {}
\end{verbatim}

@@

Here is the code of the corresponding composed pNet:
\noteEMin{Il faudrait (mais a-t-on le temps?): faire un dessin de la
  composition au niveau BIP, et au niveau pNet. Un resultat brut, vaguement
  correct, montre 37 OTs, modulo les petites questions de mon mail de
  mercredi midi...

  Ecrire les proprietes qu'il doit verifier... ? }

@@

Fig. \ref{schema:ArchFailure:OA-TimerV4X2} shows the open automata of
this composed architecture.
Note that e..g. the open transition labelled ``$fail(*,tt)$'' stands for
``$fail(b01,b02)$'', where  $b01'$ can be arbitrarily true or
false. As in the previous case, this is because we have not use here
the Max Progress hypothesis, otherwise all the ``*'' in the figure
should be replaced by ``$ff$''.

  \begin{figure}[h]
  \centering
  \includegraphics[width=\columnwidth]{TimerV4X2}
  \caption{The open Automaton of the composed system}
  \label{schema:ArchFailure:OA-TimerV4X2}
\end{figure}

>>>>

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.9\columnwidth]{FailureTimerPNET-v4}
  \caption{The open pNET encoding the Failure Monitor architecture (\fig{schema:ArchFailure:BIP}) without the Max Progress priority model}
  \label{fig:FailureTimer:pNet}
%% %
%%   \noteSBin{The updated encoding exports more primed variables in the
%%     actions, \eg $tick(z,t,z',t')$ and $fail(zone, zone', b1)$ instead
%%     of $tick(z,t,z')$ and $fail(zone, b1)$, and SVs, \eg $\dots \land
%%     (t' = t) \land (zone' = zone)$ in the guard of SV0. Shall we
%%     update the figure to include these or, rather, explain that the
%%     figure is cleaned-out for readability?}
%
\end{figure}

@@

The encoding of the interaction model is based on its representation
as a set of connectors. \noteSB{%
%
%%   A legitimate question is whether we actually need to do this,
%%   instead of listing explicitly the individual interactions, then
%%   associating a synchronisation vector to each one of them?  One of
%%   the factors is the encoding pNets$\rightarrow$SAT: is it better to
%%   have more or fewer SVs if the price is additional Boolean variables?
%%   The other factor might be the facility of maximal progress
%%   encoding\mdash TBC.  Traceability.  Anything else?\newline
%%   \hl{UPDATE: Maximal progress encoding requires this!}
%% %
} Indeed, as illustrated by the Failure Monitor architecture in
\fig{schema:ArchFailure:BIP}, each connector can define several
allowed interactions, depending on its hierarchical structure and the
use of synchrons and triggers.

@@

\addSB{In the general encoding, the global actions of the
  synchronisation vectors should be fresh action names.  However, for
  the properties considered in our case study, we only need to exhibit
  the actions of the holes.  Furthermore, each synchronisation vector
  involves at most one such action.  This allows us to reuse them as
  the global actions of the synchronisation vectors, simplifying the
  translation of properties to be verified (\cf \app{properties}).}

@@

\noteEM{Dans le paragraph precedent, tu resume le cas des connecteurs sans data, dans les prochains, tu formalise avec data, mais pour le lecteur ``casual'', le changement n'est pas evident... }
Below, we assume that, as in \fig{schema:ArchFailure:BIP}, the
interaction model is defined by a set of connectors, annotated with
Boolean guards and with update assignments.  In particular, we assume
that the guards and update assignments are well-defined for any
interaction allowed by the connector.  For example, the choice
$\NameCtrl.\PortFail\ ?\ \NameCtrl.\mathit{zone} : \topinterval$ in
the update assignement
$
\NameTimer.\mathit{z} := \NameTimer.\mathit{z}
\cap \bigl(
(
\NameCtrl.\PortFail\ ?\ \NameCtrl.\mathit{zone} : \topinterval
)
+ \NameTimer.\mathit{t} \bigr)
$
associated to the connector {\NameTimer.\PortTick \trigsynch
  (\PortFail \trigsynch \NameCtrl.\PortFail)} in
\fig{schema:ArchFailure:BIP} ensures that the assignment is
well-defined independently of whether {\NameCtrl.\PortFail}
participates or not.  Let us denote by $\gamma \subset \ac(P_A)
\times \guards{V_A} \times \assigns{V_A}$ the set of connectors in the
architecture $A$ and by $P_x$ the set of ports involved in the
connector $x \in \gamma$.  Then, \addSB{the interaction model defined by $\gamma$ is 
  $\Gamma = \setdef{(a,g,u)}{(x,g,u) \in \gamma, a \in \intsem{x}}$
  and the encoding of $\Gamma$ is} $\nopri[\Gamma] \bydef{=}
\setdef{\nopri[x,g,u]}{(x,g,u) \in \gamma}$, where $\nopri[x,g,u]$ is the synchronisation vector defined by
%
\begin{multline}
  \label{eq:nopri:connector}
  \nopri[x,g,u] \bydef{=}
  %% \Tuple{p\bigl(\export[A](p),{{\color{blue}\export[A]'(p),}}b_p\bigr)}^{p \in P_x}
  \Tuple{\Setdef{p\bigl(\export[A](p),\export[A]'(p),b_p\bigr)}{p \in P_x \cap P}}^{P \in P_C^{C \in \cC} \cup \partition}
  \rightarrow
  \alpha_p
  \\
  \left[
    \bigwedge R(\tau(x))\bigl[\nicefrac{b_p}{p}\bigr]
    \land
    \bigwedge_{(x := e_x) \in u} (?x' = e_x)
    \land
    \bigwedge_{x \in \export[A](x), x \not\in u} (?x' = x)
    \right]
  \,,
\end{multline}
%% \noteSB{Ought to group ports by components in
%%   $\Tuple{p\bigl(\export[A](p),\export[A]'(p),b_p\bigr)}^{p \in P_x}$...
%%   \newline
%%   Will probably restrict interactions to one port per component
%%   throughout the paper to simplify.
%% }
%
with $b_p$ a fresh Boolean variable, $\alpha_p$ a fresh name, $\tau :
\ac(P_A) \rightarrow \ct(P_A)$ and $R : \ct(P_A) \rightarrow
\cru(P_A)$ the two mappings defined in \app{transformations} and
illustrated in \tab{connectors}, $\bigl[\nicefrac{b_p}{p}\bigr]$ is
the substitution that replaces in the expression that preceeds it
all occurrences of $p$ by $b_p$.

@@

\bigskip
In order to encode the maximal progress priority model, one has to
introduce an additional Boolean variable, for each port, for which we
have introduced one in $\nopri[A, \partition]$.  Intuitively, the
variables introduced above determine whether the original transition
labeled by the port is executed (\true), or rather the corresponding
self-loop, introduced by the encoding (\false).  The new variables
determine whether \emph{there is an original transition labelled by
  $p$ that could be executed from the same state}, \ie[,] with $b_p$
the variable introduced above for the encoding without maximal
progress, $q \goesto{p(\true, b_p)} q'$ iff $\exists q'': q
\goesto{p(b_p)} q''$ with $b_p = \true$ in $\nopri[A, \partition]$.
Then, in the SV guard, we have to check whether \emph{all ports
  leading to $p$ in the causal tree $\tau(x)$ can be fired} (see
\app{transformations}).  If so, then $p$ must be fired, \ie
$p(\false)$ must be blocked.  %% This is pretty much straightforward, but
%% care must be taken when doing this in the general case, since causal
%% interaction trees can get pretty tricky due to $\oplus$ (see
%% \app{trees}).}
%****************************************************************
%****************************************************************

@@

   {t1---cancel(false)-->t1, s1---resume(false)-->s1},   {---:hb_B-->},
   [(:hb_B=fail(b0:sva_SV0)
     /\((cancel(false))=(cancel(b1:sva_SV1)))
     /\((resume(false))=(resume(b2:sva_SV1)))
     /\(b1:sva_SV1=b2:sva_SV1)],   {}
OT4 ------------------------------------------------------------
    <t1_s1>-----resume----><t1_s1>

@@

  In particular, the first safety property \eq{prop:reset:can}:
  \begin{equation*}
    \AG \bigl(\PortFail(true) \rightarrow \EU[]{\lnot \PortFail(true)}{\PortReset}\bigr).
  \end{equation*}
is not true, because the $OT_2$ loop on state $\ostate{11}$ stays in
state 11 where you need another $\PortFail(true)$ before enabling a
reset. This is because without priorities we can have an
activity on the dandling ``fail'' port without synchronising with the
others, even if they are available.

@@

pLTS: Timer
- Transition: t0---init-->t1, { t := 0 z := TopInt}
- Transition: t1---tick(z, t, z1)-->t1, [t<(GET_MAX z)], { t := t+1 z := z1}
- Transition: t1---cancel(true, false)-->t1, {}
- Transition: t1---cancel(true, true)-->t1, { t := 0 z := TopInt}
- Transition: t1---timeout-->t1, [t in z], { t := 0 z := TopInt}
pLTS: Control
- Transition: s0---init-->s1, { zone := minZone,maxZone}
- Transition: s1---timeout-->s1, {}
- Transition: s1---resume(false, false)-->s1, {}
- Transition: s1---fail(zone, true, false)-->s1, {}
- Transition: s1---fail(zone, true, true)-->s2, {}
- Transition: s2---fail(zone, false, false)-->s2, {}
- Transition: s2---resume(true, false)-->s2, {}
- Transition: s2---resume(true, true)-->s1, {}
- Transition: s2---timeout-->s3, {}
- Transition: s3---fail(zone, false, false)-->s3, {}
- Transition: s3---resume(false, false)-->s3, {}
- Transition: s3---reset-->s1, {}
\end{verbatim}

@@

Here is the code of the corresponding composed pNet:
\noteEMin{Il faudrait (mais a-t-on le temps?): faire un dessin de la
  composition au niveau BIP, et au niveau pNet. Un resultat brut, vaguement
  correct, montre 37 OTs, modulo les petites questions de mon mail de
  mercredi midi...}

\todoSBin{Ecrire les proprietes qu'il doit verifier... ? }

@@

\bigskip
Fig. \ref{schema:ArchFailure:OA-TimerV4X2} shows the open automata of
this composed architecture.
Note that e..g. the open transition labelled ``$fail(*,tt)$'' stands for
``$fail(b01,b02)$'', where  $b01'$ can be arbitrarily true or
false. As in the previous case, this is because we have not use here
the Max Progress hypothesis, otherwise all the ``*'' in the figure
should be replaced by ``$ff$''.

  \begin{figure}[h]
  \centering
  \includegraphics[width=0.8\columnwidth]{TimerV4X2}
  \caption{The open Automaton of the composed system}
  \label{schema:ArchFailure:OA-TimerV4X2}
\end{figure}
